---
title: "Data375Project"
author: "Ava Mizrahi"
date: "2025-04-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data sets

```{r}
data1 <- read.csv("/Users/avamizrahi/Downloads/Student_Loan_Debt.csv")

data2 <- read.csv("/Users/avamizrahi/Downloads/College_Rankings.csv")

head(data1)
head(data2)
```

Combining the 2 datasets
```{r}
data1$School <- tolower(data1$School)
data2$School <- tolower(data2$College.Name)

College_Data <- merge(data1, data2, by = "School")
head(College_Data)
College_Data <- na.omit(College_Data)
Ranking_Tuition <- College_Data

library(dplyr)
colnames(College_Data)
colnames(Ranking_Tuition) <- c(
  "School", "index", "OPE.ID", 
  "City", "State", "Zip.Code", 
  "School.Type", "LoanType", "Recipients", 
  "XLoansOriginated", "PLoansOriginated", "XDisbursements", 
  "PDisbursements", "CollegeName", "AdjustedRank", 
  "Tuition", "EnrollmentNumbers"
)
```

```{r}
Ranking_Tuition <- Ranking_Tuition %>%
  group_by(School) %>%
  summarize(
    Tuition = mean(Tuition),
    Rank = mean(AdjustedRank),
    Enrollment = mean(EnrollmentNumbers),
    Num_Loans = sum(XLoansOriginated),
    Total_Loan_amt = sum(PLoansOriginated),
    Num_Disbursements = sum(XDisbursements),
    Total_Disbursements_amt = sum(PDisbursements)
  )
colnames(Ranking_Tuition)
```
```{r}
loan_data <- College_Data
colnames(loan_data) <- c(
  "School", "index", "OPE.ID", 
  "City", "State", "Zip.Code", 
  "School.Type", "LoanType", "Recipients", 
  "XLoansOriginated", "PLoansOriginated", "XDisbursements", 
  "PDisbursements", "CollegeName", "AdjustedRank", 
  "Tuition", "EnrollmentNumbers"
)
loan_data <- loan_data %>%
  group_by(LoanType) %>%
  summarize(
    Recipients = sum(Recipients),
    Num_Loans = sum(XLoansOriginated),
    Total_Loan_amt = sum(PLoansOriginated),
    Num_Disbursements = sum(XDisbursements),
    Total_Disbursements_amt = sum(PDisbursements)
  )

#loan_data

loan_type_average_loan <- loan_data$Total_Loan_amt/loan_data$Num_Loans
#loan_type_average_loan
mean_loans <- cbind(loan_data$LoanType, as.double(loan_type_average_loan))
mean_loans

```

Summary Statistics
```{r}
summary(Ranking_Tuition)

# Summary of average loan per student
Ranking_Tuition$AvgLoanPerStudent <- Ranking_Tuition$Total_Loan_amt / Ranking_Tuition$Num_Loans
summary(Ranking_Tuition$AvgLoanPerStudent)

```


Correlation Analysis with Rank and Tuition
```{r}
cor.test(Ranking_Tuition$Rank, Ranking_Tuition$AvgLoanPerStudent)
cor.test(Ranking_Tuition$Tuition, Ranking_Tuition$AvgLoanPerStudent)
```
Rank vs. Average Loan per Student: The correlation is −0.552 (moderate negative correlation). The p-value is 2.58e-09, which is highly significant. This indicates a strong inverse linear relationship: as school rank increases (i.e., rank gets worse), the average loan per student tends to decrease.

Tuition vs. Average Loan: The correlation is 0.541 (moderate positive correlation). The p-value is 6.36e-09, which is strongly significant. This suggests that higher tuition is associated with higher average student loan debt.



Linear Regression Model
```{r}
model_rank <- lm(AvgLoanPerStudent ~ Rank, data = Ranking_Tuition)
summary(model_rank)

model_tuition <- lm(AvgLoanPerStudent ~ Tuition, data = Ranking_Tuition)
summary(model_tuition)

model_both <- lm(AvgLoanPerStudent ~ Rank + Tuition, data = Ranking_Tuition)
summary(model_both)

```
Avg Loan vs Rank: For each one-place increase in rank (e.g., from #1 to #2), the average loan per student decreases by about $70.82. The p-value for rank is < 0.000000003, indicating strong statistical significance. The model's R^2 is 0.305, meaning that 30.5% of the variation in average student loan debt is explained by school rank.

Avg Loan vs Tuition: For every 1 increase in tuition, the average loan increases by about $0.25. The p-value is < 0.000000006, showing strong significance. The R² is 0.292, indicating that tuition alone explains 29.2% of the variation in student loan burden.

Avg Loan vs Rank and Tuition: Each 1-rank increase decreases average loan per student by about 45.51 (p = 0.0007, statistically significant). Each 1 dollar increase in tuition increases average loan per student by about $0.15 (p = 0.0018, strongly significant). Together, rank and tuition explain 37.2% of the variability in average student loan burden. Rank remains the stronger factor, but tuition also makes a meaningful contribution.


Regression Models
```{r}
par(mfrow = c(2,2))   
plot(model_rank)
par(mfrow = c(2,2))
plot(model_tuition)
par(mfrow = c(2,2))
plot(model_both)
par(mfrow = c(1,1))  
```


Monte Carlo Simulation
```{r}
# rate_ mean is 4 because college tuition has increased 3%–5% annually in the U.S. rate_sd adds random variation 

simulate_loans <- function(tuition_start, rate_mean = 0.04, rate_sd = 0.01, years = 10) { 
  rates <- rnorm(years, mean = rate_mean, sd = rate_sd)
  tuitions <- tuition_start * cumprod(1 + rates)  # get compound growth
  predicted_loans <- predict(model_tuition, newdata = data.frame(Tuition = tuitions))
  return(predicted_loans[years])
}

set.seed(123)
sim_results <- replicate(10000, simulate_loans(tuition_start = 40000))

#Histogram to show the simulation
hist(sim_results, breaks = 50, main = "Projected Avg Loan After 10 Years",
     xlab = "Predicted Avg Loan")

```
If tuition continues to grow at around 4% annually, and the current relationship between tuition and loan size holds, the average loan per borrower could increase to around $16,500 in 10 years. Each bar represents how often a certain average loan value occurred across 10,000 simulated futures, incorporating year-to-year tuition variability.



95% Confidence Intervals for Mean Loan
```{r}
mean_loan <- mean(Ranking_Tuition$AvgLoanPerStudent)
se_loan <- sd(Ranking_Tuition$AvgLoanPerStudent) / sqrt(nrow(Ranking_Tuition))
lower_bound <- mean_loan - 1.96 * se_loan
upper_bound <- mean_loan + 1.96 * se_loan
c(lower_bound, upper_bound)

```



```{r}
predict_loan <- function(rank, tuition) {
  new_data <- data.frame(Rank = rank, Tuition = tuition)
  prediction <- predict(model_both, newdata = new_data, interval = "prediction", level = 0.95)
  return(prediction)
}

# example
predict_loan(rank = 30, tuition = 60000)
```



Loan Efficiency Analysis


```{r}
Ranking_Tuition$LoanWaste <- Ranking_Tuition$Total_Loan_amt - Ranking_Tuition$Total_Disbursements_amt
summary(Ranking_Tuition$LoanWaste)
top_waste <- Ranking_Tuition[order(-Ranking_Tuition$LoanWaste), ]
head(top_waste[, c("School", "LoanWaste")], 10)

```































































































